Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-11 paper2.qmd
section:appendix; process:write,figures

This section uses the structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] to articulate a theoretical model that captures the core principles of Thurstone's theory. The model also incorporates various assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. Finally, the section employs Bayesian statistical methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data. See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section.

@@

@@

## Empirical research under causal inference {#sec-appendix-B .appendix appendix-style=plain}

<!-- 1. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, the *causal inference* framework is designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in understanding and defining causal effects.

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)$. Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$. Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable. To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts. Notably, the approach extends the ACEs for categorical and continuous variables $T$, ensuring broad applicability across different causal scenarios.

@@

For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and instead rely on observational data, they can still estimate the ACE provided a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ to the outcome $Y$  without opening any new ones. Under these conditions, the ACE can be estimated from associational quantities as $\tau = E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014].

@@

<!-- 3. the SCMs -->
Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks. The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X,Z,Y\}$, a set of *exogenous* variables $E=\{e_{X},e_{Z},e_{Y}\}$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence of the variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

::: {#fig-scm_bb1 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

<!-- ![Two unconnected nodes](images/png/sdag_bb1.png){#fig-sdag_bb1 fig-align="center" fig-pos="H" width=35%} -->

@@

::: {#fig-scm_bb2 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Y & := f_{Y}(X,e_{Y}) \\
  e_{X} & \dsep e_{Y}
\end{aligned}
$$

@@

<!-- ![Two connected nodes or descendant](images/png/sdag_bb2.png){#fig-sdag_bb2 fig-align="center" fig-pos="H" width=35%} -->

@@

::: {#fig-scm_bb3 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Z} \\
  e_{Z} & \dsep e_{Y}
\end{aligned}
$$

@@

<!-- ![Chain or pipe](images/png/sdag_bb3.png){#fig-sdag_bb3 fig-align="center" fig-pos="H" width=55%} -->

@@

::: {#fig-scm_bb4 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(Z,e_{X}) \\
  Z & := f_{Z}(e_{Z}) \\
  Y & := f_{Y}(Z,e_{Y}) \\
  e_{X} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Z} \\
  e_{Z} & \dsep e_{Y}
\end{aligned}
$$

@@

<!-- ![Fork](images/png/sdag_bb4.png){#fig-sdag_bb4 fig-align="center" fig-pos="H" width=55%} -->

@@

::: {#fig-scm_bb5 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  Z & := f_{Z}(X,Y,e_{Z}) \\
  Y & := f_{Y}(e_{Y}) \\
  e_{X} & \dsep e_{Y} \\
  e_{X} & \dsep e_{Z} \\
  e_{Z} & \dsep e_{Y}
\end{aligned}
$$

@@

<!-- ![Collider or inmorality](images/png/sdag_bb5.png){#fig-sdag_bb5 fig-align="center" fig-pos="H" width=55%} -->

@@

<!-- 5. Implications of the basic building block of the SCMS and DAGs -->
A careful examination of the figures highlights the assumptions underlying these building blocks. Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related. Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$. Consequently, $Y$ is considered a *descendant* of $X$. Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X$ influences $Z$, and $Z$ influences $Y$. In this configuration, $X$ is a parent node of $Z$, which is a parent node of $Y$. This structure creates a *directed path* between $X$ and $Y$. Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two. Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$. Here, $Z$ is a parent node that *confounds* the relationship between $X$ and $Y$. Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X$ and $Y$ are concurrent causes of $Z$. In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$ (an *inmorality*). Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.

@@

<!-- ![Standard representation](images/png/sdag_example1.png){#fig-sdag_example1 fig-align="center" width=55%} -->

@@

## The flow of association and causation in DAGs {#sec-appendix-D .appendix appendix-style=plain}

<!-- 1. The general probabilistic description of data -->

<!-- Maximum a posteriori (MAP) estimates are the set of parameters that maximizes the probability of $\theta$ given the data. By Bayes theorem the MAP estimate can be obtained by maximizing the following conditional distribution: -->

<!-- $$ -->
<!-- P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})} -->
<!-- $$ -->
<!-- where $P(\theta)$ is the prior distribution for the parameters $\theta$ and $P(X_{1}, X_{2}, \dots, X_{n})$ is the probability of the data averaged over all parameters. Since the denominator is independent of $\theta$, the Bayesian estimator is obtained by maximizing $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$ with respect to $\theta$. This indicates that when researchers model association, they generally model the joint probability of the data given the parameters $\theta$. -->

<!-- This join probability can be further re-expressed using the chain rule of probability as $P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )$. However, from this description, it is easy to see how quickly the parametrization becomes intractable as the number of observed variables $n$ grows [@Neal_2020]. To deal with this complexity, researchers typically simplify the modeling process by assuming local (in)dependencies between variables.  -->

<!-- 2. But what does the building blocks mean? -->
<!-- The five building blocks help to express joint distributions very efficiently, by formalizing local (in)dependencies into (probabilistic) statistical models [@Neal_2020]. Notably, this formalization rest on two key assumptions: the local Markov and causal edges assumptions. The local Markov assumption states that given its parents in the DAG, a node is independent of all its non-descendants. This lead to the Bayesian Network factorization. However, this assumption only give us information about independencias in the distribution that the DAG implies, we need also to know about the dependencies. The causal edges assumption states that in a DAG, every parent is a direct cause of all its children. In this sense this assumption is indicating that parents and their children are dependent, and more precisely, they are causally dependent [@Neal_2020].  -->

<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->

@@

<!-- [The advantages of SCMs and DAGs to identify causal quantities are realized after researchers examine the causal and probabilistic assumptions underlying these building blocks.]{style="color:red;"} -->

<!-- The main assumption we’ve seen that relates the graph to the distribution -->
<!-- is the Markov assumption. The Markov assumption tells us if variables are -->
<!-- d-separated in the graph 𝐺 , then they are independent in the distribution -->
<!-- 𝑃 (Theorem 3.1): -->
<!-- Maybe we can detect independencies in the data and then use that -->
<!-- to infer the causal graph. However, going from independencies in the -->
<!-- distribution 𝑃 to d-separations in the graph 𝐺 isn’t something that the -->
<!-- Markov assumption gives us (see Equation 3.20 above). Rather, we need -->
<!-- the converse of the Markov assumption. This is known as the faithfulness -->
<!-- assumption. -->
<!-- Assumption 11.1 (Faithfulness) -->

<!-- 2. What is the local markov assumption? and where does it lead? -->
<!-- 2.1 Bayesian network factorization -->

<!-- 3. What is the causal edges assumption? and where does it lead? -->

<!-- 4. Combined they are known as the minimality assumption? -->

@@

<!-- Causal mechanism as a result of markov and minimality assumption -->
<!-- Modularity assumption also -->

<!-- 4. But how do we move from targets of inference to estimates -->
<!-- But how can we move from heuristics and vague mathematic notation, to probabilistic causal analysis. -->

>>>>

This section uses the structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] to articulate a theoretical model that captures the core principles of Thurstone's theory. The model also incorporates various assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. Finally, the section employs Bayesian inference methods to transform these theoretical and practical elements into a statistical model that facilitates the analysis of pairwise comparison data. See @sec-appendix for an overview of the statistical and causal inference concepts required for the development of this section.

@@

<!-- 2. What are the benefits of Bayesian inference?-->
Bayesian inference procedures offer three key advantages. First, they are well-suited to handling complex and overparameterized models, enabling researchers to estimate models where the number of parameters exceeds the number of observations for estimation [@Baker_1998; @Kim_et_al_1999]. Second, they allow researchers to incorporate prior information, which helps constrain parameters within specified bounds. This capability addresses challenges such as non-convergence or improper parameter estimation that often arise in complex models when analyzed with frequentist methods [@Martin_et_al_1975; @Seaman_et_al_2011]. Finally, Bayesian methods are particularly effective at drawing inferences from small sample sizes, where relying on the asymptotic properties of frequentist approaches may not be justified [@Baldwin_et_al_2013; @Lambert_et_al_2006; @Depaoli_2014].

@@

## Identification under causal inference {#sec-appendix-B .appendix appendix-style=plain}

<!-- 1. What is causal inference? -->
Unlike classical statistical modeling, which focuses primarily on summarizing data and inferring associations, the *causal inference* framework is designed to identify causes and estimate their effects using data [@Shaughnessy_et_al_2010; @Neal_2020]. The framework uses rigorous mathematical techniques to address the *fundamental problem of causality* [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This problem revolves around the question, "What would have happened 'in the world' under different circumstances?" This question introduces the concept of counterfactuals, which are instrumental in defining and identifying causal effects.

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes: $\tau_{i} = Y_{i}|do(T_{i}=1) - Y_{i}|do(T_{i}=2)$. Here, $do(T_{i}=t)$ represents the intervention operator, $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=1$, and $Y_{i}|do(T_{i}=1)$ represents the potential outcome under intervention $T_{i}=2$. Note that if a student is assigned to intervention $T_{i}=1$, the potential outcome under $T_{i}=2$ becomes a counterfactual, as it is no longer observed nor observable. To address the challenge of unobserved counterfactuals, the approach extends the ICE to the *average causal effect* (ACE): $\tau = E[\tau_{i}] = E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)]$, representing the average difference between observed potential outcomes and their counterfactual counterparts.

@@

For instance, if researchers are unable to conduct the randomized experiments described in @sec-appendix-A and instead rely on observational data, they can still estimate the ACE provided a variable $X$, such as the socio-economic status of the school, blocks all non-causal paths between the teaching method $T$ to the outcome $Y$  without opening any new ones. Under these conditions, the ACE can be estimated from associational quantities as $\tau =$ $E[Y_{i}|do(T_{i}=1)]- E[Y_{i}|do(T_{i}=2)] =$ $E_{X}\left[ E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X] \right]$, where $E_{X}[\cdot]$ represents the marginal expected value over $X$ [@Morgan_et_al_2014]. Notably, the approach extends the ACE for a continuous variable $T$ as $\tau =$ $E[Y_{i}|do(T_{i}=t)] =$ $d E_{X}\left[ E[Y_{i}| T_{i}=t, X]\right]/ dt$, ensuring broad applicability across different causal scenarios [@Neal_2020, pp. 45].

@@

<!-- 3. the SCMs -->
Figures [-@fig-dags_scms1], [-@fig-dags_scms2], [-@fig-dags_scms3], [-@fig-dags_scms4], and [-@fig-dags_scms5] display these five fundamental building blocks. The left panels of the figures show the formal mathematical models, represented by the SCMs, defined in terms of a set of *endogenous* variables $V=\{X_{1},X_{2},X_{3}\}$, a set of *exogenous* variables $E=\{e_{X1},e_{X2},e_{X3}\}$, and a set of functions $F=\{f_{X1},f_{X2},f_{X3}\}$ [@Pearl_2009; @Cinelli_et_al_2020; @Neal_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence of the variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

::: {#fig-scm_bb1 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \dsep e_{X3}
\end{aligned}
$$

@@

@@

::: {#fig-scm_bb2 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{3} & := f_{X3}(X_{1},e_{X3}) \\
  e_{X1} & \dsep e_{X3}
\end{aligned}
$$

@@

@@

::: {#fig-scm_bb3 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \dsep e_{X2} \\
  e_{X1} & \dsep e_{X3} \\
  e_{X2} & \dsep e_{X3}
\end{aligned}
$$

@@

@@

::: {#fig-scm_bb4 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X_{1} & := f_{X1}(X_{2},e_{X1}) \\
  X_{2} & := f_{X2}(e_{X2}) \\
  X_{3} & := f_{X3}(X_{2},e_{X3}) \\
  e_{X1} & \dsep e_{X2} \\
  e_{X1} & \dsep e_{X3} \\
  e_{X2} & \dsep e_{X3}
\end{aligned}
$$

@@

@@

::: {#fig-scm_bb5 fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  X_{1} & := f_{X1}(e_{X1}) \\
  X_{2} & := f_{X2}(X_{1},X_{3},e_{X2}) \\
  X_{3} & := f_{X3}(e_{X3}) \\
  e_{X1} & \dsep e_{X2} \\
  e_{X1} & \dsep e_{X3} \\
  e_{X2} & \dsep e_{X3}
\end{aligned}
$$

@@

@@

<!-- 5. Implications of the basic building block of the SCMS and DAGs -->
A careful examination of the figures highlights the assumptions underlying these building blocks. Figures [-@fig-scm_bb1] and [-@fig-mdag_bb1] depict two unconnected nodes, representing a scenario where variables $X_{1}$ and $X_{3}$ are not causally related. Figures [-@fig-scm_bb2] and [-@fig-mdag_bb2] illustrate two connected nodes, representing a scenario where a *parent* node $X_{1}$ exerts a causal influence on a *child* node $X_{3}$. In this setup, $X_{3}$ is considered a *descendant* of $X_{1}$. Additionally, $X_{1}$ and $X_{3}$ are described as *adjacent* because an edge directly connects them. Figures [-@fig-scm_bb3] and [-@fig-mdag_bb3] depict a *chain*, where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, which is a parent node of $X_{3}$. This structure creates a *directed path* between $X_{1}$ and $X_{3}$. Consequently, $X_{1}$ is an *ancestor* of $X_{3}$, and $X_{2}$ fully *mediates* the relationship between the two. Figures [-@fig-scm_bb4] and [-@fig-mdag_bb4] illustrate a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. Here, $X_{2}$ is a parent node that *confounds* the relationship between $X_{1}$ and $X_{3}$. Finally, Figures [-@fig-scm_bb5] and [-@fig-mdag_bb5] show a *collider*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$ (an *inmorality*). Notably, in all SCMs, the errors are assumed to be independent of each other and from all other variables in the graph, as evidenced by the pairwise relations $e_{X1} \dsep e_{X2}$, $e_{X1} \dsep e_{X3}$, and $e_{X2} \dsep e_{X3}$.

@@

@@

## The probabilistic implications of DAGs {#sec-appendix-D .appendix appendix-style=plain}

<!-- 1. How do we use Bayesian inference? -->
After completing the identification process outlined in @sec-appendix-B, researchers estimate the ACE using Bayesian inference methods. Bayesian inference calculates the probability of a set of parameters $\theta$, enabling researchers to use their distributions to represent the ACE. The approach begins by defining two probability distributions: the likelihood of the data, $P(X_{1}, X_{2}, \dots, X_{n} | \theta)$, and the prior distribution, $P(\theta)$ [@Everitt_et_al_2010]. Here, $X_{n}$ represents a random variable, and for simplicity, $\theta$ denotes a parameter space of dimension one. After observing empirical data, researchers update these priors to posterior distributions using Bayes' rule [@Jeffreys_1998]:

$$
P(\theta| X_{1}, X_{2}, \dots, X_{n}) = \frac{P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)}{P(X_{1}, X_{2}, \dots, X_{n})}
$$ {#eq-bayes_rule}

Researchers can further simplify the posterior updating process in @eq-bayes_rule into two steps: integrating new empirical data, which is defined by the likelihood, with the parameter update, which is determined by the priors, as demonstrated in equation @eq-prop_rule. This simplification is possible because the denominator on the right-hand side of @eq-bayes_rule acts as a normalizing constant, independent of $\theta$.

$$
P(\theta| X_{1}, X_{2}, \dots, X_{n}) \propto P(X_{1}, X_{2}, \dots, X_{n} | \theta) \cdot P(\theta)
$$ {#eq-prop_rule}

<!-- check: -->
<!-- https://en.wikipedia.org/wiki/Maximum_a_posteriori_estimation -->
<!-- https://en.wikipedia.org/wiki/Bayesian_inference -->

<!-- 2. But what is the problem? -->
Temporarily setting aside the definition of prior distributions, it is important to note that the posterior updating process depends on the assumptions underlying the likelihood of the data. However, as the number of random variables, $n$, increases, this joint distribution becomes quickly intractable [@Neal_2020]. This is evident from @eq-like_chain, where the likelihood distribution is re-expressed in terms of chained conditional distributions.

$$
P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | X_{i-1}, \dots, X_{1}, \theta )
$$ {#eq-like_chain}

<!-- 3. How do DAGs can help? -->
To address the complexity of the likelihood, researchers simplify the distribution by assuming that variables exhibit specific local (in)dependencies. These assumptions improve model tractability and streamline the estimation process. Researchers formalize these local (in)dependencies using directed acyclic graphs (DAGs). DAGs represent the probabilistic implications of these assumptions by relying on three key principles: the local Markov assumption, the minimality assumption, and the causal edges assumption. The *local Markov assumption* defines the statistical independencies implied by a DAG. It states that a variable is independent of all its non-descendants, given its parents [@Neal_2020]. The *minimality assumption* identifies the statistical dependencies implied by a DAG. It asserts that every pair of adjacent variables are dependent [@Neal_2020]. Finally, the *causal edges assumption* establishes causal relationships between parents and their children in a DAG by stating that every parent is a direct cause of their children [@Neal_2020]. Figure @fig-ACflow illustrates the flow of association and causation in graphs.

<!-- 4. What are the implications of this? -->
This lead to the *Bayesian Network factorization* of @eq-like_chain.

$$
P(X_{1}, X_{2}, \dots, X_{n}|\theta) = P(X_{1}|\theta) \prod^{n}_{i=2} P(X_{i} | pa(X_{i}), \theta )
$$ {#eq-net_factor}

@@

@@
